---
title: "widyr: Widen, process, and re-tidy a dataset"
author: "David Robinson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


This package wraps the pattern of un-tidying data into a wide matrix, performing some processing, then turning it back into a tidy form. This is useful for several mathematical operations such as co-occurence counts, correlations, or clustering that are best done on a wide matrix.

## Towards a precise definition of "wide" data

The term "wide data" has gone out of fashion as being "imprecise" [(Wickham 2014)](http://vita.had.co.nz/papers/tidy-data.pdf)), but I think with a proper definition the term could be entirely meaningful and useful.

A **wide** dataset is one or more matrices where:

* Each row is one **item**
* Each column is one **feature**
* Each value is one **observation**
* Each matrix is one **variable**

When would you want data to be wide rather than tidy? Notable examples include classification, clustering, correlation, factorization, or other operations that can take advantage of a matrix structure. In general, when you want to **compare between items** rather than compare between variables, this is a useful structure.

The widyr package is based on the observation that during a tidy data analysis, you often want data to be wide only *temporarily*, before returning to a tidy structure for visualization and further analysis. widyr makes this easy through a set of `pairwise_` functions.

## Example: gapminder

Consider the gapminder dataset in the [gapminder package](https://CRAN.R-project.org/package=gapminder).

```{r}
library(dplyr)
library(gapminder)

gapminder
```

This tidy format (one-row-per-country-per-year) is very useful for grouping, summarizing, and filtering operations. But if we want to *compare pairs of countries* (for example, to find countries that are similar to each other), we would have to reshape this dataset. Note that here, country is the **item**, while year is the **feature** column.

#### Pairwise operations

The widyr package offers `pairwise_` functions that operate on pairs of items within data. An example is `pairwise_dist`:

```{r}
library(widyr)

gapminder %>%
  pairwise_dist(country, year, lifeExp)
```

In a single step, this finds the Euclidean distance between the `lifeExp` value in each pair of countries, matching pairs based on year. We could find the closest pairs of countries overall with `arrange()`:

```{r}
gapminder %>%
  pairwise_dist(country, year, lifeExp) %>%
  arrange(distance)
```

Notice that this includes duplicates (Germany/Belgium and Belgium/Germany). To avoid those (the upper triangle of the distance matrix), use `upper = FALSE`:

```{r}
gapminder %>%
  pairwise_dist(country, year, lifeExp, upper = FALSE) %>%
  arrange(distance)
```

In some analyses, we may be interested in correlation rather than distance of pairs. For this we would use `pairwise_cor`:

```{r}
gapminder %>%
  pairwise_cor(country, year, lifeExp, upper = FALSE, sort = TRUE)
```
